\plaintitle

### 2016-09-21 Paper/proceedings.tex
Jeff's edits and comments.

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Regression by Eye: Estimating Trends in Bivariate Visualizations}
\def\plainauthor{Removed for Review}
\def\emptyauthor{}
\def\plainkeywords{Information Visualization, Graphical Perception, Regression}
\def\plaingeneralterms{Visualization}

@@

\begin{abstract}
Observing trends and predicting future values are common tasks for viewers of bivariate data visualizations. As many charts do not explicitly include trend lines or related statistical summaries, viewers often visually estimate trends directly from a plot. How reliable are the inferences viewers draw when performing such \emph{regression by eye}? Do particular visualization designs or data features bias trend perception? We present a series of crowdsourced experiments that assess the accuracy of trends estimated using regression by eye across a variety of bivariate visualizations, and examine potential sources of bias in these estimations. We find that viewers accurately estimate trends in many standard visualizations of bivariate data, but that both visual features (e.g., ``within the bar'' bias) and data features (e.g., the presence of outliers) can result in visual estimates that systematically diverge from standard least-squares regression models. In most cases regression by eye performs similarly to ordinary least squares regression, but there are edge cases where perceptual and statistical regression systematically diverge. These results suggest that, while regression by eye can be a reliable way of estimating trends, in some common cases these estimates systematically differ from statistical methods of regression.
\end{abstract}

\category{H.5.m.}{Information Interfaces and Presentation
  (e.g. HCI)}{Miscellaneous}

@@

An oft-cited example of the power of data visualization is Anscombe's quartet~\cite{anscombe1973graphs}, a set of four bivariate datasets with nearly identical summary statistics, but with qualitatively different patterns when drawn as four scatter plots (Fig. \ref{fig:anscombe}). This example relies on the fact that people have the ability to \emph{perceptually} estimate \emph{statistical} quantities of interest. Visualization users regularly perform statistical tasks, including model selection, identification of outliers, and estimation of summary statistics, entirely through visual inspection. Recent work has examined the accuracy of these visual inspections for means in scatter plots \cite{gleicher2013perception} and time series data \cite{correll2012comparing}, as well as speculated on the affordances of visualizations in general for supporting this sort of visual estimation of summary statistics \cite{szafir2016four}.

@@

Even if designers include all relevant information, the audience may lack the statistical expertise to interpret these values, or may be misled if the data violate modeling assumptions. As a further complication, the form of visual encoding may influence viewers' inferences. For instance, viewers may be more likely to consider trends with line charts, and to compare individual values with bar charts \cite{zacks1999bars}. Visual design choices can also introduce bias, such as the visual asymmetry of bars causing ``within-the-bar'' bias \cite{newman2012bar}. Designers would benefit from guidance regarding how accurately viewers make trend estimations by eye, and to what degree different visualization types might bias these estimations.

@@

In this work, we describe a series of crowdsourced experiments on the visual estimation of trends in common bivariate visualizations such as scatter plots, area charts, and line graphs. We present the results of three studies investigating estimation of trend slope, trend intercept, and the effect of outliers. We find that, while in most cases viewers accurately estimate trends, area charts introduce systematic underestimation of trend intercept, and that viewers give low weights to extreme values when estimating trends. These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments. Designers should therefore make an informed decision between designing for regression by eye and the explicit annotation of statistical regression information. Regression by eye affords quick estimations and flexibility in model selections, but is subject to perceptual biases and inaccuracies in estimation. Explicit annotation affords accuracy, but constrains the model space, and adds visual complexity to plots.

@@

Scatter plots are a standard means of visualizing bivariate data, with a multitude of design parameters that affect their suitability for aggregate tasks~\cite{cleveland1984many}. Prior work has confirmed that viewers can make use of scatter plots to perform prediction tasks (which tacitly rely on estimation of trend) in ways that are robust to both noise~\cite{harvey1997effects} and problem frame~\cite{lewandowsky2011popular}. However, the heuristics used when performing these prediction tasks (such as the anchor-and-adjust method~\cite{bolger1993context}) can introduce biases.

Similarly, visual design choices (for instance, the decision to encode class membership with color or with shape) can also impact performance at aggregate tasks~\cite{gleicher2013perception,lewandowsky1989discriminating}. The aspect ratio of graphs can also bias judgments about trends~\cite{beattie2002impact}\,---\,wider aspect ratios can cause viewers to underestimate the severity of effects, when compared to narrower charts. Another bias in prediction tasks is the ``within-the-bar'' bias \cite{newman2012bar}: for visually asymmetric visualizations such as bar charts, points contained within the visual area of the bar glyph are perceived as likelier than those outside the glyph.

Recent work in the visualization community has focused on the perception of correlation in scatter plots. Rensink et al.~\cite{rensink2010perception} show that viewers can estimate correlation with some accuracy in scatter plots. Harrison et al.~\cite{harrison2014ranking} extend this finding to other visualization types, although a re-analysis by Kay \& Heer~\cite{kay2016beyond} indicates that, for many of the more esoteric bivariate visualizations, performance at this task is poor. Estimation of correlation can also be biased by the choice of scales~\cite{cleveland1982variables}\,---\,the whitespace and aspect ratio changes introduced by expanding the scale of the axes can cause overestimation of correlation.

Our task of trend estimation combines elements of both prediction and correlation estimation tasks. As with prediction, there is not necessarily an unambiguous correct estimation of trend (different modeling and regression methods can produce different trend lines). As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.~\cite{harrison2014ranking}) likely relies on a set of visual proxies. That is, while viewers are unlikely to be directly estimation correlation \emph{per se}, they are estimating correlation through the perception of visual features such as the envelope of the points, point dispersion, or some other visual feature highly correlated with the statistic of interest.

Prior work in visualization has not directly addressed the capabilities of regression by eye, instead assuming as a given that visual estimates of trend are sufficiently accurate. Work in the perception of summary statistics lends credences to this assumption, but we believe that trend estimation can be biased through conscious or unconscious design choices. For instance, within-the-bar bias may result in underestimation of trends in bar and area charts, and factors that affect perception of correlation (such as aspect ratio and whitespace) may result in under- or overweighting of outliers.

\section{General Methods}

@@

In order to assess the ability of visualization viewers to estimate trends, we conducted a series of three crowdsourced experiments on Amazon's Mechanical Turk platform. We designed these experiments to establish a performance baseline for regression by eye, and to examine potential sources of bias for these estimates. Crowdsourcing is a valuable tool for graphical perception experiments, as it affords flexibilities and scales that would be difficult for in-person experiments~\cite{heer2010crowdsourcing}. Crowdsourced graphical perception experiments produce results that are largely in keeping with prior, lab-based work~\cite{heer2010crowdsourcing, talbot2014four}. In this section, we describe experimental design aspects shared across all experiments.

@@

We limited the Mechanical Turk participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%. For validation, certain trials contained the actual (OLS) trend line. We excluded the results of participants with poor performance on these validation questions. Across the three reported experiments and pilots, we performed 7 such exclusions. We recruited additional participants to replace these excluded subjects. Based on timings from internal piloting, we paid each participant \$2 for their participation, for a target rate of \$8/hour.

@@

We investigated three types of common bivariate visualizations: scatter plots, line graphs, and area charts (see Fig. \ref{fig:types}). In each experiment, we presented participants with a visualization and asked them to interactively adjust trend lines to best fit the data.  Participants responded using a slider without tick marks, in order to limit anchoring effects~\cite{matejka2016effect}. Moving the slider adjusted a purple trend line by modifying its slope (in experiments 1 and 3) or its y-intercept (in experiment 2). Fig. \ref{fig:experiment} shows an example experimental task. After moving the slider, participants had to confirm their choice of trend line. The primary dependent measure is accuracy (e.g., the difference between the subject-specified slope and the slope of the series if the residuals were removed). We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand. We avoided more expressive designs (such as freeform drawing or other elicitation techniques) in order to constrain the model space and to reduce the impact of factors such as accuracy of motor movements or heterogeneity of input devices.

@@

For each point set, we created a set of residual values, sampled evenly from a discretized Gaussian. The bandwidth of this Gaussian controls both goodness of fit and dispersion of points (see Fig. \ref{fig:sigmas}). We permuted this set of residuals, and then applied them to points along a target trend. As heteroskedasticity introduced through permutation could alter the actual trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend. We reused these residuals across all different trend types (linear, quadratic, or trigonometric). Fig. \ref{fig:trendtypes} shows these different trend types on the same set of residuals. Except where noted, we selected trend lines that were centered in the image: that is, for a horizontal data extent of $[0,1]$, $f(0.5) = 0.5$. In all experiments, we desired control over the direction of the trend. For linear fits, this is the slope of the trend line. For quadratic fits, this was the curvature, as controlled by the coefficient of the second degree term. For trigonometric fits, this was the amplitude of a cosine function (with negative amplitude corresponding to negative slopes, and positive to positive slopes). These terms create similar relationships, such that a value of $0$ is the line $f(x)=0.5$, a value of $1$ goes from $f(0)=0$ to $f(1)=1$, and $-1$ has the inverse relationship.

@@

We designed our first experiment to examine how accurate participants were at estimating the magnitude (slope, amplitude, or curvature) of trends in bivariate visualizations. We examined three types of bivariate visualizations: scatter plots, line graphs, and area chart (with the filled area below the line).In addition to linear trends, we examined more complex relationships such as quadratic and trigonometric functions.

We presented participants with a series of bivariate visualizations, who adjusted a slider to fit the perceived trend in the points. The slider parameterized one of three types of trend: linear, quadratic, or trigonometric. For each stimulus, participants adjusted a slider that controlled the slope of a rendered trend line. In the case of quadratic trends, this slope was the curvature; for trigonometric fits, the positive/negative amplitude.

@@

We had three hypotheses for the first experiment:
\begin{enumerate}
	\item \textbf{As the bandwidth of the residuals increased, accuracy would decrease}. Increasing the bandwidth of the residuals results in a lower correlation coefficient and higher perceived noise in the bivariate data. Prior work indicates that these related measures correspond to decreased accuracy for aggregate tasks in bivariate visualizations~\cite{albers2014task, harrison2014ranking}.
	\item \textbf{More complex relationships would result in lower accuracy}. Quadratic and trigonometric relationships are visually more complex than linear relationships, and often require more complex statistical methods to analyze. We anticipated that estimation of these less familiar relationships would therefore be more difficult than the linear case.
	\item \textbf{Estimations would be unbiased}. That is, there would be no systematic over- or underestimation of trends.
\end{enumerate}

@@

We performed a three-way analysis of covariance (ANCOVA) of the effect of residual bandwidth, graph type and trend type on absolute error in estimation of trend lines. We included participant ids and the actual slope of the trend line as covariates. In this experiment, the primary measure of performance was the interquartile mean (or midmean). The interquartile mean discards points in the first and fourth quartiles before averaging, trimming the tails of distributions. Cleveland \& McGill~\cite{cleveland1984graphical} used the interquartile mean to provide a more robust measure of central tendency for responses from graphical perception studies, where participant error can create long tailed distributions of error. This is especially the case for crowdsourced studies, where correcting for data quality issues is perceived as more difficult than in laboratory settings~\cite{heer2010crowdsourcing}.

@@

Our results fail to support our second hypothesis: \textbf{there was no statistically significant difference in estimation accuracy among linear, quadratic, or trigonometric trends}. Fit type had only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and a post-hoc tests (using Tukey's Honest Significant Difference) did not identify any significant pairwise-interactions. The interquartile mean of absolute  error of linear fits was $17\%$ of the actual slope, in the middle of the $16\%$ error for quadratic fits, and $18\%$ for trignometric fits, providing evidence that the relative unfamiliarity of non-linear trends does not have a notable impact on performance. Fig. \ref{fig:exp1types} illustrates this result.

Our results support our third hypothesis: \textbf{there was no statistically significant bias in estimations}. Participants saw an equal number of stimuli with positive and negative trend lines. If estimates of these trends were unbiased, we would expect the average signed error to be close to zero. The average signed error was $0.0008$, far less than the fidelity of the slider used to input guesses ($\Delta=0.01$). A Student's T-test failed to support the hypothesis that $mu_{error}\ne0$ ($t(4590) = 0.39$, $p=0.70$).

In general, accuracy at estimation of slope was high, an interquartile mean of $0.06$ for slope values that ranged from $\{-1,1\}$, corresponding to an accuracy of $17\%$ of the actual slope.

@@

We hypothesized that this bias would likewise occur for regression tasks when using visually asymmetric visualizations such as area charts. The slope estimation task in the previous experiment would not capture this bias, as there is no method for participants to indicate a \emph{uniform} underestimation in trends; decreasing the slope would cause underestimation at the beginning of the plot but not the end, and vice versa. We therefore designed this experiment to elicit estimates of the y-intercepts of trends. A within-the-bar bias would then appear as systematic underestimation of intercept in area charts.

@@

Our results support our first hypothesis: \textbf{participants systematically underestimated the intercept of trends in area charts, but not in scatter plots or line graphs}. We observed a significant main effect of graph type on signed error ($F(2,2431)=27$, $p<0.001$). A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatter plots and line graphs. The interquartile mean of the signed error of estimations made with area charts was an underestimation of $-0.02$, compared to an interquartile mean of $0.002$ for the other two conditions. This underestimation corresponds to unsigned errors more than twice as large in area charts (interquartile mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for line graphs and scatter plots). Figure \ref{fig:exp2types} illustrates this result.

\section{Experiment 3. Estimation involving Outliers}

@@

OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit. Extreme outliers violate this assumption, and can result in trend lines that are substantially different from those produced by more robust methods. Visual inspection can accurately identify certain classes of outliers \cite{albers2014task}. Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends. However, cognitive biases~\cite{tversky1975judgment} such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values with continue) can impact how viewers categorize and utilize outliers in prediction tasks ~\cite{campbell2009anchoring, ji2001culture}. That is, regression by eye may not uniformly weight outliers, depending on their position in the plot.

This experiment was largely identical to experiment 1 above, except we designated 0, 5, 10, or 15 points at the very beginning, first third, or end of the series as outliers, and placed them randomly in the top or bottom 10\% of the visual area of the visualization (whichever was farthest from the trend line). We then calculated the outlier-sensitive line of best fit, as well as the intersection between this new trend line and the original, outlier-less trend line. Fig. \ref{fig:outliers} shows examples of this process. As with experiment 1, the participants controlled the slope of a rendered trend line with a slider. However, rather than being placed such that $f(0.5)=0.5$, we offset the trend line to the intersection of the robust (without outliers) and non-robust (including outliers) trend line. This allowed participants to express both types of fit with the same slider interaction, while affording estimations beyond a simple interpolation of both trends. This expressiveness is only afforded by linear fits of, and so we did not consider quadratic or trigonometric trends in this experiment.

Similar to the previous experiments, participants saw one of each of the three graph types, slope, and Gaussian residual bandwidth. The four outlier quantities $\{0,5,10,15\}$ was an additional factor. To maintain a manageable number of stimuli (piloting showed evidence of fatigue for more than 100 stimuli per trial), both the sign of the trend line (positive or negative), and the location of the outliers (beginning, first third, or end) were random factors, with each level apportioned to half and one third of the stimuli respectively. This resulted in 96 total stimuli, with an additional 4 validation stimuli otherwise excluded from analysis, in line with prior experiments.

@@

We had three hypotheses for the third experiment:
\begin{enumerate}
	\item \textbf{Participant estimation would be closer to a robust trend line ignoring outliers, than the non-robust OLS trend line.} That is, we assumed that in general, participants would ignore or downweight outliers when performing regression by eye.
	\item \textbf{As the number of outliers increased, estimations would be closer to the non-robust OLS fit.} We speculated on the existence of a ``tipping point'' of outlier density, beyond which participants would avoid robust fits, and interpolate between the robust and OLS fits.
	\item \textbf{Outliers at the end of the chart would result in estimations closer to the OLS fit than outliers in other locations on the plot.} Prior work on cognitive biases in predictions suggests that viewers may give more credence to more recent outliers (as indicative as new anchors points, or of an emerging ``streak'')~\cite{croson2005gambler}. We therefore believed that these points would be more heavily weighted, and the resulting estimation closer to the outlier-sensitive OLS line, than outliers ``earlier'' in the plot.
\end{enumerate}

@@

In many cases, designers do not explicitly encode regression information. In other cases, viewers may not have the statistical or graphical expertise to interpret such information, even when it is supplied. Yet, our results point to regression by eye as a robust and accurate method for estimating trends in bivariate data. Participants from a variety of backgrounds and levels of self-reported graphical and statistical expertise were capable of estimating both the slope and intercept of both linear and non-linear trends. That is, viewers of visualization are largely trustworthy when estimating the relationship between two variables in a plot.

However, this general accuracy of trend estimation is not universal. Area charts are visually asymmetrical: the area below the line is filled in, and the area above it is not. We found that this asymmetry creates a within the bar bias: a systematic underestimation of the intercept of trends, due to the perceived higher likelihood of points in the filled-in area. Designers hoping to rely on regression by eye should avoid such asymmetries in their bivariate visualizations.

>>>>

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Regression by Eye: Estimating Trends\\in Bivariate Visualizations}
\def\plainauthor{Removed for Review}
\def\emptyauthor{}
\def\plainkeywords{Information Visualization, Graphical Perception, Regression}
\def\plaingeneralterms{Visualization}

@@

\begin{abstract}
Observing trends and predicting future values are common tasks for viewers of bivariate data visualizations. As many charts do not explicitly include trend lines or related statistical summaries, viewers often visually estimate trends directly from a plot. How reliable are the inferences viewers draw when performing such \emph{regression by eye}? Do particular visualization designs or data features bias trend perception? We present a series of crowdsourced experiments that assess the accuracy of trends estimated using regression by eye across a variety of bivariate visualizations, and examine potential sources of bias in these estimations. We find that viewers accurately estimate trends in many standard visualizations of bivariate data, but that both visual features (e.g., ``within the bar'' bias) and data features (e.g., the presence of outliers) can result in visual estimates that systematically diverge from standard least-squares regression models.
% In most cases regression by eye performs similarly to ordinary least squares regression, but there are edge cases where perceptual and statistical regression systematically diverge. These results suggest that, while regression by eye can be a reliable way of estimating trends, in some common cases these estimates systematically differ from statistical methods of regression.
% JEFF: These last two lines largely re-state the preceding sentence. Maybe add a new line to the abstract that summarizes implications for visualization designers or tool makers.
\end{abstract}

\category{H.5.2}{Information Interfaces and Presentation
  (e.g., HCI)}{Evaluation/Methodology}

@@

An oft-cited example of the power of data visualization is Anscombe's quartet~\cite{anscombe1973graphs}: a set of four bivariate datasets with identical summary statistics, but with qualitatively different patterns when drawn as four scatter plots (Fig. \ref{fig:anscombe}). This example relies on the fact that people have the ability to \emph{perceptually} estimate \emph{statistical} quantities of interest. Visualization users regularly perform statistical tasks\,---\,including model selection, identification of outliers, and estimation of summary statistics\,---\,entirely through visual inspection. Recent work examines the accuracy of visual estimation of means in scatter plots \cite{gleicher2013perception} and time series data \cite{correll2012comparing}, and speculates on the affordances of visualizations in general for supporting visual estimation of summary statistics \cite{szafir2016four}.

@@

Even if designers include modeling information, the audience may lack the statistical expertise to interpret these values, or may be misled if the data violate modeling assumptions. As a further complication, the form of visual encoding may influence viewers' inferences. For instance, viewers may be more likely to consider trends with line charts, and to compare individual values with bar charts \cite{zacks1999bars}. Visual design choices can also introduce bias, such as the visual asymmetry of bars causing ``within-the-bar'' bias \cite{newman2012bar}. Designers would benefit from guidance regarding how accurately viewers make trend estimations by eye, and to what degree different visualization types might bias these estimations.

@@

In this work, we describe a series of crowdsourced experiments on the visual estimation of trends in common bivariate visualizations such as scatter plots, area charts, and line graphs. We present the results of three studies investigating estimation of trend slope, trend intercept, and the effect of outliers. We find that, while in most cases viewers accurately estimate trends, area charts introduce systematic under-estimation of trend intercept, and that viewers give low weights to extreme values when estimating trends. These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments.

Designers should therefore make an informed decision between designing for regression by eye and the explicit annotation of statistical regression information. Regression by eye affords quick estimations and flexibility in model selections, but is subject to perceptual biases and inaccuracies in estimation. Explicit annotation affords accuracy, but constrains the model space, and adds visual complexity to plots.

@@

Scatter plots are a standard means of visualizing bivariate data, with a multitude of design parameters that affect their suitability for aggregate tasks~\cite{cleveland1984many}. Prior work has confirmed that viewers can make use of scatter plots to perform prediction tasks (which tacitly rely on trend estimation) in ways that are robust to both noise~\cite{harvey1997effects} and problem frame~\cite{lewandowsky2011popular}. However, the heuristics used to perform these prediction tasks (such as the anchor-and-adjust method~\cite{bolger1993context}) can introduce biases.

Similarly, visual design choices (for instance, the decision to encode class membership with color or with shape) can also impact performance at aggregate tasks~\cite{gleicher2013perception,lewandowsky1989discriminating}. The aspect ratio of graphs can also bias judgments about trends~\cite{beattie2002impact}, as wider aspect ratios can cause viewers to underestimate the severity of effects, when compared to narrower charts. Another bias in prediction tasks is the ``within-the-bar'' bias \cite{newman2012bar}: for visually asymmetric visualizations such as bar charts, points contained within the visual area of the bar glyph are perceived as likelier than those outside the glyph.

Recent work in the visualization community has focused on the perception of correlation in scatter plots. Rensink et al.~\cite{rensink2010perception} show that viewers can estimate correlation with some accuracy in scatter plots. Harrison et al.~\cite{harrison2014ranking} extend this finding to other visualization types, and a re-analysis by Kay \& Heer~\cite{kay2016beyond} indicates that, for many of the more esoteric bivariate visualizations, performance at this task is poor. Estimation of correlation can also be biased by the choice of axis scales~\cite{cleveland1982variables}: the whitespace and aspect ratio changes introduced by expanding the scale of the axes can cause over-estimation of correlation.

Our task of trend estimation combines elements of both prediction and correlation estimation tasks. As with prediction, there is not necessarily an unambiguous correct estimation of trend (different modeling and regression methods can produce different trend lines). As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.~\cite{harrison2014ranking}) likely relies on a set of visual proxies. That is, while viewers are unlikely to directly estimate correlation \emph{per se}, they are estimating correlation through the perception of visual features such as the envelope of the points, point dispersion, or some other visual feature(s) highly correlated with the statistic of interest.

Prior work in visualization has not directly addressed the capabilities of regression by eye, instead assuming as a given that visual estimates of trend are sufficiently accurate. Work in the perception of summary statistics lends credences to this assumption, but we believe that trend estimation can be biased through conscious or unconscious design choices. For instance, within-the-bar bias may result in under-estimation of trends in bar and area charts, and factors that affect perception of correlation (such as aspect ratio and whitespace) may result in up- or down-weighting of outliers.

\section{General Experimental Methods}

@@

In order to assess the ability of visualization viewers to estimate trends, we conducted a series of three crowdsourced experiments on Amazon's Mechanical Turk. We designed these experiments to establish a performance baseline for regression by eye, and to examine potential sources of bias. Crowdsourced graphical perception experiments have been found to produce results that are largely in keeping with prior, lab-based work~\cite{heer2010crowdsourcing, talbot2014four}. In this section, we describe experimental design aspects shared across all experiments.

@@

We limited the Mechanical Turk participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%. For validation, certain trials contained the actual (OLS) trend line, which subjects then needed to simply match. \jeff{But what if subjects honestly disagreed? Did we do this in the presence of outliers, for example?} We excluded the results of participants with poor performance on these validation questions. \jeff{What level of error constitutes poor performance?} Across the three reported experiments and pilots, we performed 7 such exclusions. We recruited additional participants to replace these excluded subjects. Based on timings from internal piloting, we paid each participant \$2 for their participation, for a target rate of \$8/hour.

@@

We investigated three types of common bivariate visualizations: scatter plots, line graphs, and area charts (see Fig. \ref{fig:types}). In each experiment, we presented participants with a visualization and asked them to interactively adjust trend lines to best fit the data.  Participants responded using a slider without tick marks, in order to limit anchoring effects~\cite{matejka2016effect}. Moving the slider adjusted a purple trend line by modifying its slope (in experiments 1 and 3) or its y-intercept (in experiment 2). Fig. \ref{fig:experiment} shows an example experimental task. After moving the slider, participants had to confirm their choice of trend line. The primary dependent measure is accuracy (e.g., the difference between the subject-specified slope and the slope of the series if the residuals were removed). We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand. We avoided more expressive designs (such as freeform drawing) in order to constrain the model space and to reduce the impact of factors such as noise from motor movements or heterogeneous input devices.

@@

For each point set, we created a set of residual values, sampled evenly from a discretized Gaussian. \jeff{In what way was it discretized?} The bandwidth of this Gaussian controls both goodness of fit and dispersion of points (see Fig. \ref{fig:sigmas}). We permuted this set of residuals, and then applied them to points along a target trend. As heteroskedasticity introduced through permutation could alter the actual trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend. We reused these residuals across all different trend types (linear, quadratic, or trigonometric). Fig. \ref{fig:trendtypes} shows these different trend types on the same set of residuals.

Except where noted, we selected trend lines that were centered in the image: that is, for a horizontal data extent of $[0,1]$, $f(0.5) = 0.5$. In all experiments, we desired control over the direction of the trend. For linear fits, this is the slope of the trend line. For quadratic fits, this was the curvature, as controlled by the coefficient of the second degree term. For trigonometric fits, this was the amplitude of a cosine function (with negative amplitude corresponding to negative slopes). These terms create similar relationships, such that a value of $0$ is the line $f(x)=0.5$, a value of $1$ goes from $f(0)=0$ to $f(1)=1$, and $-1$ has the inverse relationship.

@@

We designed our first experiment to examine how accurate participants were at estimating the magnitude (slope, amplitude, or curvature) of trends in bivariate visualizations. We examined three types of bivariate visualizations: scatter plots, line graphs, and area charts (with the filled area below the line). In addition to linear trends, we examined more complex relationships such as quadratic and trigonometric functions.

We presented participants with a series of bivariate visualizations, who adjusted a slider to fit the perceived trend. The slider parameterized one of three types of trend: linear, quadratic, or trigonometric. For each stimulus, participants adjusted a slider that controlled the slope of a rendered trend line. In the case of quadratic trends, this slope was the curvature; for trigonometric fits, the positive/negative amplitude.

@@

We had three hypotheses for the first experiment:
\begin{enumerate}
	\item \textbf{As the bandwidth of the residuals increased, accuracy would decrease}. Increasing the bandwidth of the residuals results in a lower correlation coefficient and higher perceived noise in the bivariate data. Prior work indicates that these related measures correspond to decreased accuracy for aggregate tasks in bivariate visualizations~\cite{albers2014task, harrison2014ranking}.
	\item \textbf{More complex relationships would result in lower accuracy}. Quadratic and trigonometric relationships are visually more complex than linear relationships, and often require more complex statistical methods to analyze. We anticipated that estimation of these less familiar relationships would therefore be more difficult than the linear case.
	\item \textbf{Estimations would be unbiased}. That is, there would be no systematic over- or under-estimation of trends.
\end{enumerate}

@@

We performed a three-way analysis of covariance (ANCOVA) of the effect of residual bandwidth, graph type and trend type on absolute error in estimation of trend lines. We included participant ids and the actual slope of the trend line as covariates. \jeff{Add a sentence here explicitly defining our error measure.}

In this experiment, the primary measure of performance was the interquartile mean (or midmean) of absolute error. The interquartile mean discards points in the first and fourth quartiles before averaging, trimming the tails of distributions. Cleveland \& McGill~\cite{cleveland1984graphical} used the interquartile mean to provide a more robust measure of central tendency for responses from graphical perception studies, where participant error can create long tailed distributions of error. This is especially the case for crowdsourced studies, where correcting for data quality issues is perceived as more difficult than in laboratory settings~\cite{heer2010crowdsourcing}.

@@

Our results fail to support our second hypothesis: \textbf{there was no statistically significant difference in estimation accuracy among linear, quadratic, or trigonometric trends}. Fit type had only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and post-hoc tests (using Tukey's Honest Significant Difference) did not identify any significant pairwise interactions. The interquartile mean of absolute error of linear fits was $17\%$ of the actual slope, in between a $16\%$ error for quadratic fits, and $18\%$ for trignometric fits (see Fig. \ref{fig:exp1types}). This result suggests that the relative unfamiliarity of non-linear trends does not have a notable impact on performance.

Our results support our third hypothesis: \textbf{there was no statistically significant bias in estimations}. Participants saw a balanced set of positive and negative trends. If estimates of these trends were unbiased, we would expect the average signed error to be close to zero. The average signed error was $0.0008$, far less than the fidelity of the slider used to input guesses ($\Delta=0.01$). A Student's T-test failed to support the hypothesis that $mu_{error}\ne0$ ($t(4590) = 0.39$, $p=0.70$).

In general, the accuracy of slope estimation was high, with an interquartile mean of $0.06$ for slope values ranging from $\{-1,1\}$, corresponding to an accuracy of $17\%$ of the actual slope. \jeff{We use the term ``accuracy'' for two different measurements here, which is confusing. Also, a widow line.}

@@

We hypothesized that this bias would likewise occur for regression tasks when using visually asymmetric visualizations such as area charts. The slope estimation task in the previous experiment would not capture this bias, as there is no method for participants to indicate a \emph{uniform} under-estimation in trends; decreasing the slope would cause under-estimation at the beginning of the plot but not the end, and vice versa. We therefore designed this experiment to elicit estimates of the y-intercepts of trends. A within-the-bar bias would then appear as systematic under-estimation of intercept in area charts.

@@

Our results support our first hypothesis: \textbf{participants systematically underestimated the intercept of trends in area charts, but not in scatter plots or line graphs}. We observed a significant main effect of graph type on signed error ($F(2,2431)=27$, $p<0.001$). A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatter plots and line graphs. The interquartile mean of the signed error of estimations made with area charts was an under-estimation of $-0.02$, compared to an interquartile mean of $0.002$ for the other two conditions. This under-estimation corresponds to unsigned errors more than twice as large in area charts (interquartile mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for line graphs and scatter plots). Figure \ref{fig:exp2types} illustrates this result.

\section{Experiment 3: Estimation Involving Outliers}

@@

OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit. Extreme outliers violate this assumption, and can result in trend lines that are substantially different from those produced by more robust methods. Visual inspection can accurately identify certain classes of outliers~\cite{albers2014task}. Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends. However, cognitive biases~\cite{tversky1975judgment} such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values will continue) can impact how viewers categorize and utilize outliers in prediction tasks~\cite{campbell2009anchoring, ji2001culture}. In other words, regression by eye may not uniformly weight outliers, depending on their position in the plot.

This experiment was largely identical to Experiment 1, except we designated 0, 5, 10, or 15 points at the very beginning, first third, or end of the series as outliers. We vertically positioned outliers within the top or bottom 10\% of the visual area of the visualization (whichever was farthest from the trend line), and added random jitter. We then calculated the outlier-sensitive line of best fit, as well as the intersection between this new trend line and the original, outlier-less trend line. Figure \ref{fig:outliers} shows examples of this process.

As in Experiment 1, the participants controlled the slope of a rendered trend line with a slider. However, rather than being placed such that $f(0.5)=0.5$, we offset the trend line to the intersection of the robust (without outliers) and non-robust (including outliers) trend line. This allowed participants to express both types of fit with the same slider interaction, while affording estimations beyond a simple interpolation of both trends. We limited the stimuli to linear fits only, excluding quadratic and trigonometric trends from this experiment.

Similar to the previous experiments, participants saw one of each of the three graph types, slope, and Gaussian residual bandwidth. \jeff{Awkward grammar for previous sentence.} The four outlier quantities $\{0,5,10,15\}$ was an additional factor. To maintain a manageable number of stimuli (piloting showed evidence of fatigue for more than 100 stimuli per trial), both the sign of the trend line (positive or negative), and the location of the outliers (beginning, first third, or end) were random factors, with each level apportioned to half and one third of the stimuli respectively. This resulted in 96 total stimuli, with an additional 4 validation stimuli otherwise excluded from analysis, in line with prior experiments.

@@

We had three hypotheses for the third experiment:
\begin{enumerate}
	\item \textbf{Participant estimation would hew closer to a robust trend line ignoring outliers, rather than the OLS trend line.} We assumed that in general, participants would ignore or downweight outliers when performing regression by eye.
	\item \textbf{As the number of outliers increased, estimations would be closer to the non-robust OLS fit.} We speculated on the existence of a ``tipping point'' of outlier density, beyond which participants would avoid robust fits, and interpolate between the robust and OLS fits.
	\item \textbf{Outliers at the end of the chart would result in estimations closer to the OLS fit than outliers in other locations on the plot.} Prior work on cognitive biases in predictions suggests that viewers may give more credence to more recent outliers (as indicative as new anchor points, or of an emerging ``streak'')~\cite{croson2005gambler}. We therefore believed that these points would be more heavily weighted, and the resulting estimation closer to the outlier-sensitive OLS line, than outliers ``earlier'' in the plot.
\end{enumerate}

@@

In many cases, designers do not explicitly encode regression information. In other cases, viewers may not have the statistical or graphical expertise to interpret such information, even when it is supplied. Yet, our results point to regression by eye as a robust and reasonably accurate method for estimating trends in bivariate data. Participants from a variety of backgrounds and levels of self-reported graphical and statistical expertise were capable of estimating both the slope and intercept of both linear and non-linear trends. That is, viewers of visualization are largely trustworthy when estimating the relationship between two variables in a plot.
